### 2025-07-10 paper/quarto_aaai/paper.tex
results section updated

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this as ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this result
holds theoretically:\footnote{For the proof, see the supplementary
  appendix.}

@@

\noindent Proposition \ref{prp-mtblty} assumes independent features,
which raises a concern about the effect of protecting immutable features
in the presence of proxies. We address it in
Section~\ref{sec-discussion}.

\subsection{Example: Prediction of Credit
Defaults}\label{example-prediction-of-credit-defaults}

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
\(y\in\{1=\text{default},2=\text{no default}\}\) with two input
features: (1) the subjects' \emph{age}, which we define as immutable,
and (2) the subjects' existing level of \emph{debt}, which we define as
mutable. We have simulated this scenario using synthetic data with
independent \emph{age} and \emph{debt} features, and Gaussian
class-conditional densities in Figure~\ref{fig-poc}. The four panels
show the outcomes for different training procedures using the same model
architecture (a linear classifier). In panels (a) and (c) we have
trained the models conventionally, while in panels (b) and (d) we have
applied CT.

In all four cases, all counterfactuals (stars) are valid---they have
crossed the decision boundary (green) into the target class---but their
quality differs. In panel (a), they are not plausible: they do not
comply with the distribution of the factuals in \(y^+\) to the point
where they form a clearly discernible cluster. In panel (b), they are
highly plausible, meeting the first objective of Def.
\ref{def-explainability}. In panel (c), the CEs involve substantial
reductions in \emph{debt} for younger applicants. By comparison,
counterfactual paths are shorter on average in panel (d) where we have
protected the immutable \emph{age} as described in
Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
to \emph{age}, recommendations with respect to \emph{debt} are much more
homogenous and do not unfairly punish younger individuals. These
counterfactuals are also plausible with respect to the mutable feature.
Thus, we consider the model in panel (d) as the most explainable
according to Def. \ref{def-explainability}.

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). Thus, we focus primarily on the plausibility
and cost of faithfully generated counterfactuals at test time. To
measure the cost, we follow the standard convention of using distances
(\(\ell_1\)-norm) between factuals and counterfactuals as a proxy. For
plausibility, we assess how similar CEs are to the observed samples in
the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
metric used by \citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

In addition to cost and plausibility, we compute other standard metrics
to evaluate counterfactuals such as validity and redundancy. Finally, we
also assess the predictive performance of models using standard metrics,
including robust accuracy estimated on adversarially perturbed data
using FGSM \citep{goodfellow2014explaining}.

We run the experiments with three gradient-based generators:
\emph{Generic} of \citet{wachter2017counterfactual} as a simple baseline
approach, \emph{REVISE} \citep{joshi2019realistic} that aims to generate
plausible counterfactuals using a surrogate Variational Autoencoder
(VAE), and \emph{ECCo}---the generator of \citet{altmeyer2024faithful}
but without the conformal prediction component---as a method that
directly targets both faithfulness and plausibility of the
counterfactuals.

@@

Table~\ref{tbl-main} presents our main empirical findings. For all
datasets except \emph{OL} and across all test settings, the average
distance of CEs from observed samples in the target class is reduced,
indicating improved plausibility. The magnitude of improvements varies.
For the simple synthetic datasets, distance reductions range from around
20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
real-world tabular datasets, improvements tend to be smaller but still
substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
for our proposed divergence metric are qualitatively similar, but
generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by the absolute outcome. Improvements for other datasets range
from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
reduction is negative, consistent with the distance-based metric.
\emph{MNIST} is the only dataset for which the two metrics disagree.
However, upon visual inspection of the image counterfactuals (available
in the supplementary appendix), we find that CT clearly improves
plausibility.

@@

\caption{\label{tbl-main}Key performance metrics across all datasets
(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
reduction in implausibility (\(\text{IP}\)) for varying degrees of the
energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
aggregated across all test specifications. \textbf{Accuracy} (columns
8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
and the baseline (BL). \textbf{Actionability} (column 12): average
reduction in costs when imposing mutability constraints reported for the
four datasets for which we could identify meaningful features to
protect.}

@@

In Section~\ref{sec-method}, we show that our method of encoding
mutability constraints decreases sensitivity to immutable features for
linear models, tilting the decision boundary in favour of mutable
features. For constraints binding at test time, this leads to shorter
counterfactual paths and hence smaller average costs (\(\ell_1\)-norm)
to individuals. To extend this to the non-linear case, we test the
effect of imposing mutability constraints for the synthetic datasets
using the same evaluation scheme as above. The final row in
Table~\ref{tbl-main} reports the average reduction in costs for CT
compared to the ``vanilla'' baseline, when imposing that either the
first or the second feature is immutable. In all cases, costs are
reduced substantially, indicating that classifiers trained with CT are
indeed more sensitive to mutable features.

>>>>

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this as ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This can be expected to result in lower overall sensitivity to
immutable features, which we investigate empirically in
Section~\ref{sec-experiments}. Under certain conditions, this result
holds theoretically:\footnote{For the proof, see the supplementary
  appendix.}

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). Thus, we focus primarily on the plausibility
and cost of faithfully generated counterfactuals at test
time\footnote{Additional metrics such as validity and redundancy are
  reported in the appendix.}. To measure the cost, we follow the
standard convention of using distances (\(\ell_1\)-norm) between
factuals and counterfactuals as a proxy. For plausibility, we assess how
similar CEs are to the observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

We also assess the predictive performance of models using standard
metrics, including robust accuracy estimated on adversarially perturbed
data using FGSM \citep{goodfellow2014explaining}.

We run experiments with three gradient-based generators: \emph{Generic}
of \citet{wachter2017counterfactual} as a simple baseline approach,
\emph{REVISE} \citep{joshi2019realistic} that aims to generate plausible
counterfactuals using a surrogate Variational Autoencoder (VAE), and
\emph{ECCo}---the generator of \citet{altmeyer2024faithful} but without
the conformal prediction component---as a method that directly targets
both faithfulness and plausibility of the counterfactuals.

Our main quantitative results for the \emph{MLP} are summarised in
Table~\ref{tbl-main}, which shows average outcomes along with
bootstrapped standard errors for key metrics. The following example
illustrates how to read Table~\ref{tbl-main}.

@@

\subsubsection{Synthetic Example: Prediction of Credit
Defaults}\label{synthetic-example-prediction-of-credit-defaults}

Figure~\ref{fig-poc} presents results for a linear classifier fitted to
\emph{LS}, which complies with the data assumptions in Proposition
\ref{prp-mtblty}. The four panels show the outcomes for different
training procedures: in panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we have applied CT. For
illustrative purposes, suppose the first feature represents \emph{debt}
(mutable) and the second feature represents \emph{age} (immutable) of
loan applicants seeking counterfactual explanations for moving to the
target class: loan provided (orange).

In all four cases, it is possible to generate valid counterfactuals
(stars) for unsuccessful applicants (blue). They cross the decision
boundary (green) into the target class, but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly discernible cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. This
difference in outcomes is quantified for the non-linear MLP in the first
two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
implausibility: it is substantial and statistically significant for
\emph{LS} across both metrics, the distance-based \(IP\) (29\%) and
divergence-based \(IP^{*}\) (55\%).

In panel (c) of Figure~\ref{fig-poc}, the CEs involve substantial
reductions in \emph{debt} for younger applicants. By comparison,
counterfactual paths are shorter on average in panel (d) where we have
protected the immutable \emph{age} as described in
Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
to \emph{age}, recommendations with respect to \emph{debt} are much more
homogenous and do not unfairly punish younger individuals. These
counterfactuals are also plausible with respect to the mutable feature,
despite requiring smaller debt reductions on average, resulting in
smaller costs to individuals. This result is quantified for the
non-linear case in column three of Table~\ref{tbl-main}, which shows the
\(\%\)-reduction in costs averaged across valid counterfactuals. Once
again, the impact of CT is statistically significant and substantial
(14\%). Thus, we consider the model in panel (d) as the most explainable
according to Def. \ref{def-explainability}. In the following, we present
the results for all remaining datasets.

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
the latter case, do not align with qualitative findings.

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as
illustrated in Figure~\ref{fig-poc}. In column three of
Table~\ref{tbl-main}, we impose mutability constraints on selected
features and compute the reduction in average costs of counterfactuals
associated with CT compared to the weak baseline: for synthetic
datasets, we always protect the first feature; for all real-world
tabular datasets we could identify and protect an \emph{age} variable;
for \emph{MNIST}, we protect the five top and bottom rows of digits.
Reductions in costs are overwhelmingly positive and significant of up to
nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
\emph{Adult} and \emph{MNIST} are not significant, FIGURE demonstrates
that CT does have the expected effect: sensitivity to protected features
as proxied by integrated gradients is drastically reduced\footnote{Additional
  findings for integrated gradients are reported in the appendix.}. In
the case of \emph{Cred}, average costs increase, likely because any
potential benefits from protecting the \emph{age} are outweighed by the
increase in costs required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
standard errors for all datasets. \textbf{Plausibility} (columns 1-2):
percentage reduction in implausibility for \(IP\) and \(IP^*\),
respectively; \textbf{Actionability} (column 3): percentage reduction in
costs with protected features. \textbf{Accuracy} (columns 4-7): test
accuracies and robust accuracies (\(\text{Acc}^*\)) for CT and the
baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated
across bootstrap samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Standard
errors for accuracy are bootstrapped from the test set.}

@@

### 2025-07-10 paper/sections/experiments.qmd
results section updated

Our key outcome of interest is improvement in explainability (Def. \ref{def-explainability}). Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals such as validity and redundancy. Finally, we also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.

@@

@@

@tbl-main presents our main empirical findings. For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*. For the vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) up to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree. However, upon visual inspection of the image counterfactuals (available in the supplementary appendix), we find that CT clearly improves plausibility.

@@

Key performance metrics across all datasets (column 1). **Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications. **Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). **Actionability** (column 12): average reduction in costs when imposing mutability constraints reported for the four datasets for which we could identify meaningful features to protect.
:::

In @sec-method, we show that our method of encoding mutability constraints decreases sensitivity to immutable features for linear models, tilting the decision boundary in favour of mutable features. For constraints binding at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals. To extend this to the non-linear case, we test the effect of imposing mutability constraints for the synthetic datasets using the same evaluation scheme as above. The final row in @tbl-main reports the average reduction in costs for CT compared to the "vanilla" baseline, when imposing that either the first or the second feature is immutable. In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.

>>>>

Our key outcome of interest is improvement in explainability (Def. \ref{def-explainability}). Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time^[Additional metrics such as validity and redundancy are reported in the appendix.]. To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

We also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.

@@

Our main quantitative results for the *MLP* are summarised in @tbl-main, which shows average outcomes along with bootstrapped standard errors for key metrics. The following example illustrates how to read @tbl-main.

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange). Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### Synthetic Example: Prediction of Credit Defaults

@fig-poc presents results for a linear classifier fitted to *LS*, which complies with the data assumptions in Proposition \ref{prp-mtblty}. The four panels show the outcomes for different training procedures: in panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT. For illustrative purposes, suppose the first feature represents *debt* (mutable) and the second feature represents *age* (immutable) of loan applicants seeking counterfactual explanations for moving to the target class: loan provided (orange).

In all four cases, it is possible to generate valid counterfactuals (stars) for unsuccessful applicants (blue). They cross the decision boundary (green) into the target class, but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $IP$ (29%) and divergence-based $IP^{*}$ (55%).

In panel (c) of @fig-poc, the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals. This result is quantified for the non-linear case in column three of @tbl-main, which shows the $\%$-reduction in costs averaged across valid counterfactuals. Once again, the impact of CT is statistically significant and substantial (14%). Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}. In the following, we present the results for all remaining datasets.

@@

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$. Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings.

<!-- TODO: Add figure for MNIST -->

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, FIGURE demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.]. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

<!-- TODO: Add figure for MNIST -->

@@

Key performance metrics and bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Standard errors for accuracy are bootstrapped from the test set.
:::

### 2025-07-10 paper/sections/method.qmd
results section updated

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

@@

`\noindent`{=latex} Proposition \ref{prp-mtblty} assumes independent features, which raises a concern about the effect of protecting immutable features in the presence of proxies. We address it in @sec-discussion.

## Example: Prediction of Credit Defaults

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable $y\in\{1=\text{default},2=\text{no default}\}$ with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable. We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architecture (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.

<!-- Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time. In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence. -->

In all four cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green) into the target class---but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange). Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

>>>>

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

@@
